---
title: "0.2Demographic patterns"
output: github_document 

---

```{r}
knitr::opts_knit$set(root.dir = "C:/Users/jc527762/OneDrive - James Cook University/PhD dissertation/Data/Chapter2_genetics")
```

# Observed and expected heterozygosity 

We used plink2 to calculate heterozygosity per sample using the following command

```{bash, echo=TRUE, eval=FALSE}
plink2 --vcf gatk.filteredv3.vcf.gz --allow-extra-chr --het --bad-freqs
```

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggpubr)
library(cowplot)
het <- read_tsv("C:/Users/jc527762/OneDrive - James Cook University/PhD dissertation/Data/Chapter2_Genetics/02.DemographicPatterns/plink2.het") 
#region.list <- factor(rep(c("CORE", "CHAUVEL","LEADING","HERON","LEADING","TORRES STRAIT","REEF HQ"), c(24, 8, 8, 8, 16, 15, 3))) 
pop.list <- factor(rep(c("Sudbury", "Tongue","Vlassof","Chauvel","Cockermouth","Heron","Keswick", 
                         "St. Bee's", "Torres", "Captive"), c(9, 7, 8, 8, 8, 8, 8, 8, 15, 3))) 
mycol=c("#FF9900","red","firebrick1","firebrick4","#3366FF","#6699FF","#66CCFF","#00BCD4","#455A64","#212121")
``` 

As part of this calculation PLINK provides the observed and expected heterozygosity (Ho, He) as well as the inbreeding coefficient **F** which is calculated as $1-\frac{H0}{He}$ 

```{r}
het2 <- het |>
  rename(sampleID = '#IID') |> 
  mutate(population = pop.list, 
         Ho = (OBS_CT-`O(HOM)`)/OBS_CT, 
         He = (OBS_CT-`E(HOM)`)/OBS_CT) |> 
  pivot_longer(cols = c(F,Ho,He), names_to = "stat", values_to = "value") |> 
  mutate(population=fct_relevel(population, c("Torres","Tongue","Vlassof","Sudbury","Cockermouth","Keswick", "St. Bee's", "Chauvel", "Heron", "Captive")))

```

```{r, fig.retina=2,fig.height=3.5,fig.width=7}


f_plot <- het2 |> 
  filter(stat=="F") |> 
  ggplot(aes(x=population,y=value)) + 
  geom_boxplot(aes(color=population),alpha=0.5) + 
  geom_point(aes(x=population,color=population)) + 
  theme_pubr() + 
  theme(legend.position = "none", 
        text = element_text(size=12) , 
        axis.text.x = element_text(size=12, angle =45, hjust =1)) +
  scale_color_manual(values = mycol) + 
  xlab("") + ylab("Inbreeding Coeffient");f_plot


het_plot <- het2 |> 
  filter(stat!="F") |> 
  ggplot(aes(x=population,y=value)) + 
  geom_boxplot(aes(fill=stat,color=stat),alpha=0.5) + 
#  geom_point(aes(x=location,color=stat), position = position_jitter(w=0.1,h=0)) + 
  theme_pubr() + 
  theme(legend.position = c(.25,.95), 
        legend.justification = c("right", "top"), 
        legend.box.just = "right", 
        legend.margin = margin(6, 6, 6, 6),
        legend.title = element_blank(), 
        text = element_text(size=12) , 
        axis.text.x = element_text(size=12, angle =45, hjust =1)) +
#  scale_color_manual(values = myCol) + 
  xlab("") + ylab("Heterozygosity"); het_plot

plot_grid(f_plot,het_plot,labels=c("A","B"))

```
In addition to filters outlined previously in section [0.1filtering_snps](https://github.com/schmidte10/Chapter2_Genetics/blob/main/01.filtering_snps.md) further filtering steps for were performed for population structure analysis (PCA, ADMIXTURE) as follows:  

1.To avoid potential errors from sequencing, we only included sites with minor allele count greater than or equal to two. 

Create a map file for _A. polyacanthus_.  

```bash 
bcftools view -H gatk.vcf.gz | cut -f 1 | uniq | awk '{print $0"\t"$0 }' > str_analysis/SCH11429.DPg90gdp3gq30.chrom-map.txt
```