---
title: "09a.PCAdapt"
output: github_document 
--- 

```bash
module load plink2
module load bcftools 
mdoule load tabix
```

convert VCF file to PLINK

```bash
bcftools view -m2 -M2 -v snps data/gatk.filtered.relaxed_studywide.bicolor.vcf.gz -Oz -o data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz 

tabix -p vcf data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz

plink2 --vcf data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz --make-bed --allow-extra-chr --out data/gatk.filtered.relaxed_studywide.bicolor.biallelic.test.vcf.gz
```
# Load libraries

```{r eval=FALSE, echo =TRUE}
library("pcadapt")
library("ggplot2")
```

# Import data

```{r eval=FALSE, echo =TRUE}
apoly_bed <- "data/gatk.filtered.relaxed_studywide.bicolor.biallelic.test.bed"
apoly_pcadapt <-  read.pcadapt(apoly_bed, type = "bed")
```

# PCAdapt

```{r eval=FALSE, echo =TRUE}
apoly_populations_kplot <- pcadapt(input = apoly_pcadapt, K=10)
pdf("outputs/apoly_populations_kplot.pdf")
plot(apoly_populations_kplot, option = "screeplot")
dev.off() 
```

A K value of 2 is most approproate, as this is the value of K after which the curve starts to flatten out more. This means we have identified the PC's that capture population structure. 

```{r eval=FALSE, echo =TRUE}
apoly_populations_pca <- pcadapt(input =apoly_pcadapt, K =3)
write.table(apoly_populations_pca$scores, "outputs/pcadapt_scores.txt")
summary(apoly_populations_pca)
```

Load file that has list of samples and populations 

Example: 

sample1   1   pop1 

sample2   1   pop1 

sample3   2   pop2 

sample4   3   pop3 

```{r eval=FALSE, echo =TRUE}
poplist.names <- read.delim("data/population_list.txt", header =FALSE)
```

# PCA

## PC1 v PC2
```{r eval =FALSE, echo =TRUE}
pdf("outputs/apoly_populations_pca1v2.pdf")
plot(apoly_populations_pca, option = "scores", i=1, j=2, pop = poplist.names$V3)
dev.off()
```

## PC1 v PC3
```{r eval =FALSE, echo =TRUE}
pdf("outputs/apoly_populations_pca1v3.pdf")
plot(apoly_populations_pca, option = "scores", i=1, j=3, pop = poplist.names$V3)
dev.off()
```

## PC2 v PC3
```{r eval =FALSE, echo =TRUE}
pdf("outputs/apoly_populations_pca2v3.pdf")
plot(apoly_populations_pca, option = "scores", i=2, j=3, pop = poplist.names$V3)
dev.off()
```

# Manhattan plot
```{r eval =FALSE, echo =TRUE}
png("outputs/pcadapt_apoly_manhattan.png")
plot(apoly_populations_pca, option= "manhattan") 
dev.off() 
```

# qqplot
```{r eval =FALSE, echo =TRUE}
png("outputs/pcadapt_apoly_qqplot.png") 
plot(apoly_populations_pca, option ="qqplot") 
dev.off()
```

# p-values of outliers
```{r eval =FALSE, echo =TRUE}
apoly_pcadapt_pvalues <- as.data.frame(apoly_populations_pca$pvalues) 
pdf("outputs/apoly_populations_pvalues.pdf")
hist(apoly_populations_pca$pvalues, xlab ="p-values", main =NULL, breaks =50, col ="skyblue") 
dev.off() 
```

# significant outliers
```{r eval =FALSE, echo =TRUE}
apoly_populations_padj <- p.adjust(apoly_populations_pca$pvalues, method ="bonferroni") 
alpha <- 0.1 #[adjust as necessary]
outliers <- which(apoly_populations_padj < alpha) 
length(outliers) 

write.table(outliers, file="outputs/apoly_populations_pcaoutliers.txt") 
```