---
title: "09.Outlier Analysis"
output: github_document 
--- 

```bash
module load plink2
module load bcftools 
mdoule load tabix
```

convert VCF file to PLINK

```bash
bcftools view -m2 -M2 -v snps data/gatk.filtered.relaxed_studywide.bicolor.vcf.gz -Oz -o data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz 

tabix -p vcf data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz

plink2 --vcf data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz --make-bed --allow-extra-chr --out data/gatk.filtered.relaxed_studywide.bicolor.biallelic.test.vcf.gz
```
# Load libraries

```{r eval=FALSE, echo =TRUE}
library("pcadapt")
library("ggplot2")
```

# Import data

```{r eval=FALSE, echo =TRUE}
apoly_bed <- "data/gatk.filtered.relaxed_studywide.bicolor.biallelic.test.bed"
apoly_pcadapt <-  read.pcadapt(apoly_bed, type = "bed")
```

# Outlier analysis
## PCAdapt

```{r eval=FALSE, echo =TRUE}
apoly_populations_kplot <- pcadapt(input = apoly_pcadapt, K=10)
pdf("outputs/apoly_populations_kplot.pdf")
plot(apoly_populations_kplot, option = "screeplot")
dev.off() 
```

A K value of 2 is most approproate, as this is the value of K after which the curve starts to flatten out more. This means we have identified the PC's that capture population structure. 

```{r eval=FALSE, echo =TRUE}
apoly_populations_pca <- pcadapt(input =apoly_pcadapt, K =3)
write.table(apoly_populations_pca$scores, "outputs/pcadapt_scores.txt")
summary(apoly_populations_pca)
```

Load file that has list of samples and populations 

Example: 

sample1   1   pop1 

sample2   1   pop1 

sample3   2   pop2 

sample4   3   pop3 

```{r eval=FALSE, echo =TRUE}
poplist.names <- read.delim("data/population_list.txt", header =FALSE)
```

## PCA

### PC1 v PC2
```{r eval =FALSE, echo =TRUE}
pdf("outputs/apoly_populations_pca1v2.pdf")
plot(apoly_populations_pca, option = "scores", i=1, j=2, pop = poplist.names$V3)
dev.off()
```

### PC1 v PC3
```{r eval =FALSE, echo =TRUE}
pdf("outputs/apoly_populations_pca1v3.pdf")
plot(apoly_populations_pca, option = "scores", i=1, j=3, pop = poplist.names$V3)
dev.off()
```

### PC2 v PC3
```{r eval =FALSE, echo =TRUE}
pdf("outputs/apoly_populations_pca2v3.pdf")
plot(apoly_populations_pca, option = "scores", i=2, j=3, pop = poplist.names$V3)
dev.off()
```

### Manhattan plot
```{r eval =FALSE, echo =TRUE}
png("outputs/pcadapt_apoly_manhattan.png")
plot(apoly_populations_pca, option= "manhattan") 
dev.off() 
```

### qqplot
```{r eval =FALSE, echo =TRUE}
png("outputs/pcadapt_apoly_qqplot.png") 
plot(apoly_populations_pca, option ="qqplot") 
dev.off()
```

### p-values of outliers
```{r eval =FALSE, echo =TRUE}
apoly_pcadapt_pvalues <- as.data.frame(apoly_populations_pca$pvalues) 
pdf("outputs/apoly_populations_pvalues.pdf")
hist(apoly_populations_pca$pvalues, xlab ="p-values", main =NULL, breaks =50, col ="skyblue") 
dev.off() 
```

### significant outliers
```{r eval =FALSE, echo =TRUE}
apoly_populations_padj <- p.adjust(apoly_populations_pca$pvalues, method ="bonferroni") 
alpha <- 0.05 #[adjust as necessary]
outliers <- which(apoly_populations_padj < alpha) 
length(outliers) 

write.table(outliers, file="outputs/apoly_populations_pcaoutliers_p05.txt") 
```

### Get outlier SNPS

Before starting make a copy of the **apoly_populations_pcaoutliers_p05.txt** file, and in the new copy removed the headers in the file, this new version should be labelled apoly_populations_pcaoutliers_p05**_no_headers.txt**. 

```{bash}
bcftools query -f '%CHROM\t%POS\n' data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz > data/all_sites.txt

awk 'NR==FNR {pos[$1]; next} ($2 in pos)' PCAdapt/outputs/apoly_populations_pcaoutliers_p05_no_headers.txt data/all_sites.txt > data/matched_sites_PCAdapt.txt

awk '{count[$2]++; chroms[$2]=chroms[$2]","$1} END {for (pos in count) if (count[pos]>1) print pos, substr(chroms[pos], 2)}' data/matched_sites_PCAdapt.txt | sort -k1,1n > data/matched_sites_PCAdapt_duplicates.txt

awk 'NR==FNR {dup[$1]; next} !($1 in dup)' data/matched_sites_PCAdapt_duplicates.txt PCAdapt/outputs/apoly_populations_pcaoutliers_p05_no_headers.txt > data/apoly_populations_pcaoutliers_p05_no_headers_no_duplicates.txt
```


## BAYPASS

### load software
```bash
module load plink
```
For this analysis use the metadata file that lists populations in the first column and individuals in the second column.  

First we will create a .ped file 
```bash 
plink --vcf data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz --recode --allow-extra-chr --out baypass/gatk.filtered.relaxed_studywide.bicolor.biallelic.plink
```

Remove first two columns 
```bash 
cut -d' ' -f3- baypass/gatk.filtered.relaxed_studywide.bicolor.biallelic.plink.ped > baypass/x.delete
```

```bash
paste -d' ' data/apoly_populations_metadata_POPIND.txt baypass/x.delete > baypass/gatk.filtered.relaxed_studywide.bicolor.biallelic.plink.ped
```

```bash
plink --file baypass/gatk.filtered.relaxed_studywide.bicolor.biallelic.plink --allow-extra-chr --freq counts --family --out baypass/apoly_populations
``` 

```bash
npop2=12 #the number of populations you have times 2 
tail -n +2 baypass/apoly_populations.frq.strat | awk '{ $9 = $8 - $7 } 1' | awk '{print $7,$9}' | tr " " "\n" | awk -v pp=${npop2} '{if (NR % pp == 0){a=a $0"";print a; a=""} else a=a $0" "}' > baypass/apoly_populations_baypass.txt 
```
## run BAYPASS 

```bash 
g_baypass singularity run /fast/tmp/containers/baypass_public-v2.41.sif g_baypass -npop 6 -gfile baypass/apoly_populations_baypass.txt -outprefix apoly_populations_baypass -nthreads 4
```

```bash 
module load tabix
bgzip -c -d ../data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz > ../data/gatk.filtered.relaxed_studywide.bioclor.biallelic.vcf

grep -v "^#" ../data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf | cut -f1-3 | awk '{print $0"\t"NR}' > ../data/apoly_populations_SNPs.txt

awk '{print $2}' PCAdapt/outputs/apoly_populations_pcaoutliers.txt > PCAdapt/outputs/apoly_populations_pcadapt_outliers_numbers.txt

awk 'FNR==NR{a[$1];next} (($4) in a)' PCAdapt/outputs/apoly_populations_pcadapt_outliers_numbers.txt data/apoly_populations_SNPs.txt   | cut -f3 > PCAdapt/outputs/pcadapt_outlierSNPIDs.txt

```

```{r echo=TRUE, eval =FALSE} 
module load R 
module show R/4.4.0 
R /fast/tmp/containers/R-4.4.0/sif R 

library(ape)
library(mvtnorm)
source("../Rscripts/baypass_utils.R")

omega <- as.matrix(read.table("apoly_populations_baypass_mat_omega.out"))

pi.beta.coef <- read.table("apoly_populations_baypass_summary_beta_params.out", header = TRUE)

bta14.data <- geno2YN("apoly_populations_baypass.txt")

simu.bta <- simulate.baypass(omega.mat = omega, nsnp = 5000, sample.size = bta14.data$NN, beta.pi = pi.beta.coef$Mean, pi.maf = 0, suffix = "btapods")

q()
```

```bash
g_baypass singularity run /fast/tmp/containers/baypass_public-v2.41.sif g_baypass -npop 6 -gfile baypass/G.btapods -outprefix G.btapods -nthreads 2
```

```{r echo=TRUE, eval =FALSE} 
R /fast/tmp/containers/R-4.4.0/sif R 

source("../Rscripts/baypass_utils.R")
library("ape")
library("corrplot")

pod.xtx <- read.table("G.btapods_summary_pi_xtx.out", header = T)
pod.thresh <- quantile(pod.xtx$M_XtX ,probs = 0.99)
pod.thresh

#9.520471

q()
```

```bash 
cat apoly_populations_baypass_summary_pi_xtx.out | awk '$4>9.520471' > baypass_outliers.txt
```
```bash 
awk 'FNR==NR{a[$1];next} (($4) in a)' baypass/baypass_outliers.txt data/apoly_populations_SNPs.txt | cut -f2 > baypass/baypass_outlierSNPIDs.txt
wc -l baypass_outlierSNPIDs.txt 
```

## Baypass 

```{bash}
awk 'NR==FNR {pos[$1]; next} ($2 in pos)' baypass/baypass_outlierSNPIDs.txt data/all_sites.txt > data/matched_sites_baypass.txt

awk '{count[$2]++; chroms[$2]=chroms[$2]","$1} END {for (pos in count) if (count[pos]>1) print pos, substr(chroms[pos], 2)}' data/matched_sites_baypass.txt | sort -k1,1n > data/matched_sites_baypass_duplicates.txt

awk 'NR==FNR {dup[$1]; next} !($1 in dup)' data/matched_sites_baypass_duplicates.txt baypass/baypass_outlierSNPIDs.txt > baypass/baypass_outlierSNPIDs_no_duplicates.txt

zcat data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz | awk 'NR==FNR {pos[$1]; next} /^#/ {print; next} ($2 in pos)' baypass/baypass_outlierSNPIDs_no_duplicates.txt - | bgzip > baypass/baypass_OutlierSNPs_filtered.vcf.gz
```


## Baypass with environmental/random variables (12)

## BaypassEnv

Before running this code remove column names and the first row that has the population names and convert the metadata file from a **.csv** file to a **.txt** file. 

Now we can run BAYPASS again and determine which SNPs are associated with summer average temperature metrics. 

```bash
g_baypass singularity run /fast/tmp/containers/baypass_public-v2.41.sif g_baypass -npop 6 -gfile apoly_populations_baypass.txt -efile metadata/env_sum.txt -scalecov -omegafile apoly_populations_baypass_mat_omega.out -outprefix apoly_populations_baypass_avg_summer -nthreads 4
```

## Baypass environment 
```{r eval =FALSE, echo =TRUE}
module load R 
module show R/4.4.0 
R /fast/tmp/containers/R-4.4.0/sif R 

library(tidyverse)
library(data.table)

covaux.snp.res.sum <- data.table::fread("baypass/env_var/apoly_populations_baypass_env_summary_betai_reg.out", header =TRUE)
covaux.snp.xtx.sum <- data.table::fread("baypass/env_var/apoly_populations_baypass_env_summary_pi_xtx.out", header =TRUE)

png("baypass_plots_env.png")
layout(matrix(1:3,3,1))
plot(covaux.snp.res.sum$`BF(dB)`, ylim=c(20,80))
abline(h =35, col ="red")
plot(covaux.snp.res.sum$Beta_is)
plot(covaux.snp.xtx.sum$XtXst, ylim =c(30, 60))
dev.off()
```

```bash 
cat baypass/env_var/apoly_populations_baypass_env_summary_betai_reg.out | awk '$7>35' > baypass/env_var/apoly_populations_baypass_env_BF35.txt
wc -l baypass/env_var/apoly_populations_baypass_env_BF35.txt
cat baypass/env_var/apoly_populations_baypass_env_BF35.txt | awk '$10>2' > baypass/env_var/apoly_populations_baypass_env_BF35eBPis2.txt
```
## Baypass environmental variables 

```bash 
awk 'FNR==NR{a[$2];next} (($4) in a)' baypass/env_var/apoly_populations_baypass_env_BF35eBPis2.txt data/apoly_populations_SNPs.txt | cut -f2 > baypass/env_var/baypass_env_outlierSNPIDs.txt

wc -l baypass/env_var/baypass_env_outlierSNPIDs.txt 

awk 'NR==FNR {pos[$1]; next} ($2 in pos)' baypass/env_var/baypass_env_outlierSNPIDs.txt data/all_sites.txt > data/matched_sites_baypass_env.txt
```

## Baypass random variables 

```{r eval =FALSE, echo =TRUE}
module load R 
module show R/4.4.0 
R /fast/tmp/containers/R-4.4.0/sif R 

library(tidyverse)
library(data.table)

covaux.snp.res.sum <- data.table::fread("baypass/random12_var/apoly_populations_baypass_ran12_summary_betai_reg.out", header =TRUE)
covaux.snp.xtx.sum <- data.table::fread("baypass/apoly_populations_baypass_ran12_summary_pi_xtx.out", header =TRUE)

png("baypass_plots_ran12.png")
layout(matrix(1:3,3,1))
plot(covaux.snp.res.sum$`BF(dB)`, ylim=c(20,80))
abline(h =35, col ="red")
plot(covaux.snp.res.sum$Beta_is)
plot(covaux.snp.xtx.sum$XtXst, ylim =c(30, 60))
dev.off()
```

```bash

cat baypass/random12_var/apoly_populations_baypass_ran12_summary_betai_reg.out | awk '$7>35' > baypass/random12_var/apoly_populations_baypass_ran12_BF35.txt
wc -l baypass/random12_var/apoly_populations_baypass_ran12_BF35.txt
cat baypass/random12_var/apoly_populations_baypass_ran12_BF35.txt | awk '$10>2' > baypass/random12_var/apoly_populations_baypass_ran12_BF35eBPis2.txt

awk 'FNR==NR{a[$2];next} (($4) in a)' baypass/random12_var/apoly_populations_baypass_ran12_BF35eBPis2.txt data/apoly_populations_SNPs.txt | cut -f2 > baypass/random12_var/baypass_ran12_outlierSNPIDs.txt

wc -l baypass/random12_var/baypass_ran12_outlierSNPIDs.txt 

awk 'NR==FNR {pos[$1]; next} ($2 in pos)' baypass/random12_var/baypass_ran12_outlierSNPIDs.txt data/all_sites.txt > data/matched_sites_baypass_ran12.txt

awk '{count[$2]++; chroms[$2]=chroms[$2]","$1} END {for (pos in count) if (count[pos]>1) print pos, substr(chroms[pos], 2)}' data/matched_sites_baypass_ran12.txt | sort -k1,1n > data/matched_sites_baypass_ran12_duplicates.txt

awk 'NR==FNR {dup[$1]; next} !($1 in dup)' data/matched_sites_baypass_ran12_duplicates.txt baypass/random12_var/baypass_ran12_outlierSNPIDs.txt > baypass/random12_var/baypass_ran12_outlierSNPIDs_no_duplicates.txt
```




```{r}
pcadapt <- scan("09.Outlier_Analysis/summary/apoly_populations_pcaoutliers_p05_no_headers_no_duplicates.txt", what = "", quiet = TRUE)
baypass <- scan("09.Outlier_Analysis/summary/baypass_outlierSNPIDs_no_duplicates.txt", what = "", quiet = TRUE)
baypass_env <- scan("09.Outlier_Analysis/summary/baypass_env_outlierSNPIDs.txt", what = "", quiet = TRUE)  
baypass_ran <- scan("09.Outlier_Analysis/summary/baypass_ran12_outlierSNPIDs_no_duplicates.txt", what = "", quiet = TRUE) 

all_outliers <- list(PCAdapt = pcadapt, Baypass = baypass, BaypassEnv = baypass_env, BaypassRan = baypass_ran)


library(UpSetR)

#pdf("All_outliers_upsetplot.pdf")
upset(
  data = fromList(all_outliers), 
  order.by = "freq", 
  empty.intersections = "on", 
  point.size = 3.5, 
  line.size = 2, 
  mainbar.y.label = "Outlier Count", 
  sets.x.label = "Total Outliers", 
  text.scale = c(1.3, 1.3, 1, 1, 2, 1.3), 
  number.angles = 30, 
  nintersects = 11
) 
#dev.off() 
```











```bash
bcftools view -v snps PCAdapt/outputs/PCAdapt_SNPs_filtered.vcf.gz -H | wc -l
grep -v "^#" baypass/baypass_OutlierSNPs_filtered.vcf.gz | wc -l
```
potentially not need 

zcat data/gatk.filtered.relaxed_studywide.bicolor.biallelic.vcf.gz | awk 'NR==FNR {pos[$1]; next} /^#/ {print; next} ($2 in pos)' PCAdapt/outputs/apoly_populations_pcaoutliers_p05_no_headers.txt - | bgzip > PCAdapt/outputs/PCAdapt_SNPs_filteredp05.vcf.gz
