---
title: "Acanthochromis polyacanthus"
output:
  github_document:
    pandoc_args: --webtex
---

```{r setup}
knitr::opts_knit$set(root.dir = "C:/Users/jc527762/OneDrive - James Cook University/PhD dissertation/Data/Local_adaptation/Chapter1_LocalAdaptation/genetic_data")
```

Load libraries 


```{r libraries, message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
library(tidyverse)
library(adegenet) 
library(ade4)
library(dartR)
library(patchwork) 
library(ape) 
library(bigsnpr) 
library(OutFLANK) 
library(vcfR)
```

```{r load_data, cache=TRUE}
setwd("C:/Users/jc527762/OneDrive - James Cook University/PhD dissertation/Data/Chapter2_Genetics")
vcf <- read.vcfR("./data/thinned_output.vcf.gz")  

genlight <- vcfR2genlight(vcf) 
save(genlight, file="./data/genlight.RData")

pop.table <- genlight@ind.names %>%
  as_tibble() %>%
  mutate(Region = str_sub(value, 1, 1),
         Population = str_sub(value, 2, 4),
         ID = str_sub(value, 5),
         across(where(is.character), as.factor), 
         Population = factor(case_when(Region == "T" ~ "DUG", 
                                TRUE ~ Population)))

genlight@pop <- pop.table$Population
genlight@ind.names <- as.character(pop.table$value)
```


```{r Allele_frequency_spectrum2, echo=TRUE, eval=FALSE}
# per population 
# need to rename  gl@pop for this to begin working 
# currently individuals are listed as pop 
for (i in levels(genlight@pop)) {
  myFreq <- glMean(genlight[genlight@pop == i])
  myFreq <- c(myFreq, 1 - myFreq)
  hist(myFreq, proba=TRUE, 
     col="cadetblue4", 
     xlab="Allele Frequencies", 
     main=paste0(i," Distribution of Allele Frequencies"), 
     nclass=20, 
     ylim = c(0,4)) ;lines(temp$x, temp$y*2, lwd=3)
}
``` 

# PCA 

#### run on hpc
```{bash, eval=FALSE, echo=TRUE}
pop.table <- gl@ind.names %>%
  as_tibble() %>%
  mutate(Region = str_sub(value, 1, 1),
         Population = str_sub(value, 2, 4),
         ID = str_sub(value, 5),
         across(where(is.character), as.factor)) 
save(pop.table, file)

gl@pop <- pop.table$Population 
gl@ind.names <- as.character(pop.table$value)

pca1 <- glPca(gl, 
              parallel = TRUE) 

save(pca1, file = "./pca1.filtered.RData")
```

Load file from hpc into R
```{r PCA, warning=FALSE} 
load("C:/Users/jc527762/OneDrive - James Cook University/PhD dissertation/Data/Chapter2_Genetics/data/genlight_pca.RData")

pca1.data <- as.data.frame(pca1$scores) 
pca1.data <- pca1.data %>% 
  mutate(value = row.names(pca1.data)) |>
  as_tibble() %>%
  mutate(Region = str_sub(value, 1, 1),
         Population = str_sub(value, 2, 4),
         ID = str_sub(value, 5),
         across(where(is.character), as.factor)) |> 
  mutate(Population = factor(case_when(Region == "T" ~ "DUG", 
                                TRUE ~ Population)))  

#--- plot ---# 
ggplot(pca1.data, aes(PC1, PC2, color = Population)) + 
  geom_point(size = 3, alpha =0.3) + 
    #geom_text(aes(label =Population), 
            #check_overlap = TRUE, 
            #nudge_x = sample(-10:10, 89, replace = TRUE), 
            #nudge_y = sample(-10:10, 89, replace = TRUE), show.legend = FALSE) + 
  theme_classic() + 
  coord_equal() + 
  #scale_color_manual(values = mycol) +
  #guides(color =guide_legend(override.aes = list(alpha =1))) + 
  theme(legend.position ='right', 
        legend.background = element_rect(color ='black'), 
        legend.title =element_blank()) + 
  xlab(paste0('PC1 (', sprintf('%0.1f%% explained var.', 100*pca1$eig[1]/sum(pca1$eig)), ")")) + 
  ylab(paste0('PC2 (', sprintf('%0.1f%% explained var.', 100*pca1$eig[2]/sum(pca1$eig)), ")"))
```

There appears to be a clear separation of fish by region, including Chauvel reef fish being on their own. According  PC1 (that explains 16.0% of the varianace), Chauvel reef fish appear to be quite different from the other populations. 

Additional pca's will be made for each population separately to see how individuals are dispersed, and to identify potential outliers 

```{r cairns_pca, warning=FALSE, eval=FALSE, echo=FALSE}
#--- split up populations ---# 
load("C:/Users/jc527762/OneDrive - James Cook University/PhD dissertation/Data/Local_adaptation/Chapter1_LocalAdaptation/genetic_data/pca.v2.filtered.cairns.RData")

pca_cairns.only.data <- as.data.frame(pca_cairns.only$scores); pca_cairns.only.data <- pca_cairns.only.data %>% 
  as_tibble() %>% 
  mutate(value =row.names(pca_cairns.only.data)) %>% 
  select(value, everything()) %>%
  mutate(Region = str_sub(value, 1, 1) ,
         Population = str_sub(value, 2, 4),
         ID = str_sub(value, 5),
         across(where(is.character), as.factor)) %>% 
  mutate(reef.names=factor(Population, levels = c("SUD","VLA","TON"), 
                      labels = c("Sudbury reef", "Vlassof cay", "Tongue reef"))); eig <- pca_cairns.only$eig 

#--- plot ---# 
ggplot(pca_cairns.only.data, aes(PC1, PC2, color = Population)) + 
  geom_point(size = 2, alpha =0.3) + 
  geom_text(aes(label =Population), 
            check_overlap = TRUE, 
            nudge_x = sample(-10:10, 89, replace = TRUE), 
            nudge_y = sample(-10:10, 89, replace = TRUE), show.legend = FALSE) + 
  theme_classic() + 
  coord_equal() + 
  scale_color_manual(values = c("#22a884","#7ad151","#fde725")) +
  guides(color =guide_legend(override.aes = list(alpha =1))) + 
  theme(legend.position =c(0.8,0.8), 
        legend.background = element_rect(color ='black'), 
        legend.title =element_blank()) + 
  xlab(paste0('PC1 (', sprintf('%0.1f%% explained var.', 100*eig[1]/sum(eig)), ")")) + 
  ylab(paste0('PC2 (', sprintf('%0.1f%% explained var.', 100*eig[2]/sum(eig)), ")"))

``` 

```{r mackay_pca, warning=FALSE, eval=FALSE, echo=FALSE}
#--- split up populations ---# 
load("C:/Users/jc527762/OneDrive - James Cook University/PhD dissertation/Data/Local_adaptation/Chapter1_LocalAdaptation/genetic_data/pca.v2.filtered.mackay.RData")

pca_mackay.only.data <- as.data.frame(pca_mackay.only$scores); pca_mackay.only.data <- pca_mackay.only.data %>% 
  as_tibble() %>% 
  mutate(value =row.names(pca_mackay.only.data)) %>% 
  select(value, everything()) %>%
  mutate(Region = str_sub(value, 1, 1) ,
         Population = str_sub(value, 2, 4),
         ID = str_sub(value, 5),
         across(where(is.character), as.factor)) %>% 
  mutate(reef.names=factor(Population, levels = c("CHA","CKM","KES"), 
                      labels = c("Cahuvel reef", "Cockermouth island", "Keswick island"))); eig <- pca_mackay.only$eig 

#--- plot ---# 
ggplot(pca_mackay.only.data, aes(PC1, PC2, color = Population)) + 
  geom_point(size = 2, alpha =0.3) + 
  geom_text(aes(label =Population), 
            check_overlap = TRUE, 
            nudge_x = sample(-10:10, 89, replace = TRUE), 
            nudge_y = sample(-10:10, 89, replace = TRUE), show.legend = FALSE) + 
  theme_classic() + 
  coord_equal() + 
  scale_color_manual(values = c("#440154","#414487","#2a788e")) +
  guides(color =guide_legend(override.aes = list(alpha =1))) + 
  theme(legend.position =c(0.5,0.8), 
        legend.background = element_rect(color ='black'), 
        legend.title =element_blank()) + 
  xlab(paste0('PC1 (', sprintf('%0.1f%% explained var.', 100*eig[1]/sum(eig)), ")")) + 
  ylab(paste0('PC2 (', sprintf('%0.1f%% explained var.', 100*eig[2]/sum(eig)), ")"))

``` 

# DAPC analysis 
Similar to example provided tutorial analysis will be conducted with both 3 and 10 axis kept 
```{r DAPC}
grp4 <- find.clusters(genlight, max.n.clust = 20, glPca = pca1, n.pca = 3, n.clust = 4)
#grp12 <- find.clusters(gl.v2.filtered, max.n.clust =20, glPca =pca1, n.pca =10, n.clust =12)

dapc4 <- dapc(genlight, grp4$grp, glPca = pca1, n.pca = 70, n.da = 3)
a_score <- optim.a.score(dapc4) # optimal is 6 PCs

#dapc12 <- dapc(gl.v2.filtered, grp12$grp, glPca =pca1, n.pca =70, n.da =3) 
#a_score <- optim.a.score(dapc6) # optimal is 7 PCs

dapc4 <- dapc(genlight, grp4$grp, glPca = pca1, n.pca =6, n.da=3)
scatter(dapc4) 

dapctbl4 <- tibble(pops = genlight@pop, inds = genlight@ind.names, gdapc4 = as.factor(dapc4$assign), xdapc4 =dapc4$ind.coord[,1], ydapc4 =dapc4$ind.coord[,2], zDAPC =dapc4$ind.coord[, 3])
head(dapctbl4)

# conserved variance
dapc4$var 

# so the variance explained by the individual discriminant aces is 
100*dapc4$var*dapc4$eig/sum(dapc4$eig) 

#--- plot ---# 
ggplot(dapctbl4, aes(xdapc4, ydapc4, color = gdapc4)) + 
  geom_point(show.legend =TRUE, 
             alpha =0.5, 
             size =7) + 
  #geom_text(aes(label = inds), 
            #show.legend =FALSE, 
            #check_overlap =TRUE, 
            #nudge_x =5) + 
  scale_color_manual(labels=c("Core","Torres","Leading","Heron/Chauvel"), values=c("#C77CFF","#00BFC4","#F8766D","#7CAE00" ))+
  theme_classic() + 
  xlab(paste0('DA1 (', sprintf('%0.1f%% explained variance',100*dapc4$var*dapc4$eig[1]/sum(dapc4$eig)),")")) + 
  ylab(paste0('DA2 (', sprintf('%0.1f%% explained variance',100*dapc4$var*dapc4$eig[2]/sum(dapc4$eig)),")"))
```

Compare DAPC without clustering - samples clustered by population
```{r DAPC_non-clustering}
#dapcPop <- dapc(gl.v2.filtered, grp =pop, glPca = pca1, n.pca =70, n.da =2)
#a_score <- optim.a.score(dapcPop) # optimal score is 10
set.seed(5)
dapcPop <- dapc(genlight, grp =pop, glPca = pca1, n.pca =23, n.da =2)
a_score <- optim.a.score(dapcPop) # optimal score is 10

dapcPoptbl <- tibble(pop =genlight@pop, 
                     inds =genlight@ind.names, 
                     gdapcPop =factor(dapcPop$assign), 
                     xdapcPop = dapcPop$ind.coord[,1], 
                     ydapcPop =dapcPop$ind.coord[,2]) 

ggplot(dapcPoptbl, aes(pop, gdapcPop)) + 
  geom_count() + 
  theme_bw() + 
  xlab("site") + 
  ylab("assigned population") + 
  scale_size_continuous(range =c(1,20)) 

g <- ggplot(dapcPoptbl, aes(xdapcPop, ydapcPop, fill=pop)) + 
  geom_point(aes(color =genlight@pop, shape = gdapcPop), 
             alpha =0.5, size=6) + 
  scale_shape_manual(values=c(0,15,22,1,2,16,17,25,3,23))+
  theme_bw() + 
  xlab(paste0('DA1 (', sprintf('%0.1f%% explained var.', 100*dapcPop$var*dapcPop$eig[1]/sum(dapcPop$eig)), ")")) +
  ylab(paste0('DA2 (', sprintf('%0.1f%% explained var.', 100*dapcPop$var*dapcPop$eig[2]/sum(dapcPop$eig)), ")")) +
  labs(shape = 'Assigned\npopulation', color = 'Site') + 
  stat_ellipse(aes(xdapcPop, ydapcPop, group =gdapcPop, color=pop), level =0.99, alpha =0.3, linetype ='dotted') + 
  theme(plot.margin =unit(c(-1,1,1,1), 'cm')) 
  #scale_color_manual(values = c("#0D47A1","#0D47A1","#0D47A1","#DA3A36","#DA3A36","#DA3A36")); g

g2 <- ggplot(dapcPoptbl, aes(xdapcPop, fill =genlight@pop, color = genlight@pop)) + 
  geom_density(alpha =0.3) + 
  theme_bw() + 
  xlab("") +
  ylab("") +
  theme(legend.position = 'none', #c(0.1, 0.8),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        plot.margin = unit(c(1, 1, -1, 1), 'cm'))

g2+g  + plot_layout(heights = c(1, 3), ncol = 1)
  
```

```{r contribtuion_of_alleles, fig.height=8, fig.width=16}
contribPopx <- loadingplot(dapcPop$var.contr, threshold = 2e-04, axis =1, lab.jitter =1, lab =genlight@loc.names, 
                           srt =90, adj=0, 
                           cex.lab =0.5) 
dapcdriverPopx <- contribPopx$var.names

freqs <- tab(genlight[,genlight@loc.names %in% contribPopx$var.names], freq = TRUE)

# by site of origin 
fr.site <- as.data.frame(freqs) %>% 
  mutate(grp = dapcPop$assign, pops = genlight@pop, inds = genlight@ind.names) %>% 
  group_by(pops) %>% 
  summarize(across(contains('NC'), list(mean))) %>% 
  pivot_longer(cols = -1, names_to = 'Locus', values_to = 'Frequency')  
fr.site$pops <- fct_relevel(fr.site$pops, c("DUG","TON","VLA","RHQ","SUD","CKM","KES","SBI","CHA","HER"))

ggplot(fr.site, aes(reorder(Locus, Frequency), pops, fill = Frequency)) + 
  geom_tile() + 
  theme_classic() + 
  coord_fixed(ratio = 2/1) + 
  theme(axis.text.x = element_blank()) + 
  xlab("DAPC group") + 
  ylab("Locus")
```

# Neighbour-joining tree 
```{r neighbour-joining tree, fig.height=8, fig.width=16}
distance <- dist(genlight)
NJ <- nj(distance)

par(mfrow=c(1,2))
mycol=c("#74ad46","#b49342","#8168cd","#4fa37c","#c86436","#d97b8d","#668fce","#a4507e","#ca5ab9","#d23f5a")
prescols=c("#CC33CC","#33CCFF","#FF9933","#660066","#33CCFF","grey77","#33CCFF","#CC0033","#CC0033","#CC0033")

plot.phylo(NJ, type = 'fan', tip.color = c(prescols)[genlight@pop], cex =0.5)
plot.phylo(NJ, type = 'fan',tip.color = c(mycol)[genlight@pop], cex =0.5)

# calculating genetic distance by 1-gl.propShared
Gdis <- 1 - gl.propShared(genlight) 
df <- as.data.frame(Gdis) %>% 
  mutate(ID =row.names(.)) %>% 
  pivot_longer(cols =1:82, values_to = 'Dist', names_to = 'ID2') %>% 
  mutate(ID2 =as.factor(ID2)) 

ggplot(df, aes(ID,ID2, fill=Dist)) + 
  geom_tile() + 
  theme_classic() + 
  coord_equal() + 
  scale_fill_distiller(palette = 'YlOrBr', direction = 1) + 
  theme(axis.text.x = element_text(angle=90))  

ggsave("nuc_tree.png", plot =nuc.tree, bg ="transparent", dpi = 300, width =8, height =4)
```

```{r fig.height=8, fig.width=16}
NJ2 <- nj(Gdis)
par(mfrow=c(1,2))
plot.phylo(NJ2, type ='fan', tip.color = c(prescols)[genlight@pop], cex =0.5)
plot.phylo(NJ2, type ='fan', tip.color = c(mycol)[genlight@pop], cex =0.5)
#plot.phylo(NJ2, type ='radial', tip.color = c(mycol)[gl.v2.filtered@pop], cex =0.5)

```


```{r diversity nucleotide diversity, eval=FALSE, echo=FALSE}
pi.all <- read.table("./diversity/SCH11429_finalgatk_10kb.windowed.pi", header =TRUE) 
#hist(pi.all$PI, br=20) 

pi.all$id <- as.numeric(row.names(pi.all))  
pi.all2 <- pi.all %>% 
  mutate(CHROM2 = as.factor(CHROM), 
         No = 1:nrow(.), 
         lanes = cut(No, breaks = seq(0, 200000, by = 20000))) %>% 
  group_by(lanes) %>% 
  mutate(id2 = seq_len(n()))
head(pi.all2)

D_poz <- pi.all2 |> 
  slice_max(order_by = PI, n =1)

D_neg <- pi.all2 |> 
  slice_min(order_by = PI, n =1)

D_sel_pi <-bind_rows(D_poz, D_neg) |> 
  droplevels()

ggplot(pi.all2, aes(x = id2, y= PI, color = CHROM)) + 
  geom_col(show.legend =FALSE) +
  geom_hline(yintercept = 0) +
  facet_wrap(~lanes, ncol = 1) +
  theme_classic() +
  xlab("") +
  ylab("PI") +
  theme(axis.text.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank())+
  geom_text(data = D_sel_pi, aes(x = id2, y = PI), label = '*', color = 'black')

```

```{r diversitytajimasD, eval=FALSE, echo=FALSE}
taj.all <- read.table("./diversity/SCH11429_finalgatk_10kb.Tajima.D", header =TRUE) 

taj.all$id <- as.numeric(row.names(taj.all)) 
head(taj.all) 

taj.all2 <- taj.all %>% 
  mutate(TajimaD =as.numeric(TajimaD), 
         CHROM2 =as.factor(CHROM), 
         No =1:nrow(.), 
         lanes =cut(No, breaks = seq(0, 200000, by = 20000))) |> 
  group_by(lanes) |> 
  mutate(id2 = seq_len(n())) 

D_poz <- taj.all2 |> 
  slice_max(order_by = TajimaD, n =1)

D_neg <- taj.all2 |> 
  slice_min(order_by = TajimaD, n =1)

D_sel_taj <-bind_rows(D_poz, D_neg) |> 
  droplevels()

head(taj.all2)

ggplot(taj.all2, aes(x=id2, y=TajimaD)) + 
  geom_col(aes(color = CHROM2), show.legend = F) +
  geom_hline(yintercept = 0) +
  facet_wrap(~lanes, ncol = 1) +
  theme_classic() +
  xlab("") +
  ylab("Tajima's D") +
  theme(axis.text.x = element_blank(), strip.background = element_blank(), strip.text.x = element_blank()) +
  geom_text(data = D_sel_taj, aes(x = id2, y = TajimaD), label = '*', color = 'black')
```




